Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-11-05 paper/manuscript.tex
Fixed typo, duplication

In the alchemical transfer approach, the standard binding free energy is obtained by a coordinate translation that brings the ligand from the solvent to the receptor binding site~\cite{wu2021alchemical} rather than by decoupling it from the system, as in the double-decoupling method. The method derives from Eq.~(\ref{eq:DG0-binding-constant-def}) by inserting the solvent degrees of freedom similarly to the double-decoupling method, except that the product $z_{R(v)} z_{L(v)}$ of partition functions of the unbound state is represented collectively as one explicit solvent system in which the ligand is placed in a region spanned by an indicator function $I^\ast(\zeta)$ identical to the one used to define the binding site region but placed at some arbitrarily large distance from the receptor:
\begin{equation}
  z_{R(v)} z_{L(v)} =
  \frac{
    \int d\vec{r}_v dx_R dx_L d\zeta I^\ast(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
  }{
    8 \pi^2 V_{\rm site} Z_{N_v}
  }
\end{equation}
where, as above, we multiplied and divided by the volume of the binding site, assuming for simplicity that the indicator function does not depend on the orientational degrees of freedom of the ligand. Combining this result with Eq.~(\ref{eq:partfunct-ratio-explsolv-1}) for the complex and canceling the common partition function of the pure solvent, leads with Eq.~(\ref{eq:DG0-binding-constant-def})  to an expression for the binding constant in terms of ratios of partition functions in explicit solvent that differ only by the location of the indicator function (in the binding site or in the solvent region):
\begin{equation}\label{eq:atm-partfunc-ratio}
  K_b = C^\circ V_{\rm site} \frac{
    \int d\vec{r}_v dx_R dx_L d\zeta I(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
  }{
    \int d\vec{r}_v dx_R dx_L d\zeta I^\ast(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
  }
\end{equation}
The idea of alchemical transfer, illustrated in Figure \ref{fig:atm-illustration}A, is to cast Eq.~(\ref{eq:atm-partfunc-ratio}) as an ansemble average by a change of coordinates noting that $I(\zeta) = I^\ast(\zeta - \vec{h})$ where $\vec{h}$ is the fixed vector distance between the binding site and solvent locations spanned by the two indicator functions. The symbol $\zeta$ represents the position and orientation of the ligand relative to the receptor, so here $\zeta - \vec{h}$ is the operation of shifting the position of the ligand by the displacement $\vec{h}$ without changing its orientation. Hence, by applying the change of variable $\zeta \rightarrow \zeta - \vec{h}$ to the integral in the numerator of Eq.~(\ref{eq:atm-partfunc-ratio}), we arrive at the following expression for the binding constant, equivalent to Eq.~(\ref{eq:DG0-binding-constant-def-average-noOmega}):
\begin{equation}\label{eq:atm-average-abfe}
K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta U} \rangle_0
\end{equation}
where
\begin{equation}\label{eq:atm-deltaU-def}
\Delta U(x) = U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) - U(x_R, x_L, \zeta, \vec{r}_v)
\end{equation}
is the perturbation energy of the method, and the ensemble average $\langle \ldots \rangle_0$ is over the unbound ensemble when the ligand is limited to the solvent region far away from the receptor. Eq.~(\ref{eq:atm-average-abfe}) instructs to simulate the unbounded system where the ligand is limited to a region in the solvent and to sample the change of potential energy $\Delta U$ resulting from transferring the ligand from the solvent to the receptor binding site by a fixed displacement of the ligand as a whole.

@@

For example, if the alchemical path you have chosen is a linear alchemical path, then $\frac{du}{d\vec{\lambda}} = u_1(\vec{q}) - u_0(\vec{q}) = u(\vec{q})$, which is the perturbation energy $u$--the potential energy difference between the initial and final states. These values are already calculated by the simulation and can be recorded easily without additional computational expense. As discussed earlier, linear paths are inefficient. However, methods such as EDS and ATM based on optimized non-linear energy interpolation alchemical energy functions $W_\lambda(u)$ that depend on atomic coordinates only through the perturbation function,~\cite{konig2021efficient,azimi2022relative} can achieve high computational efficiency by sampling and storing only perturbation energy values. The stored values can then be analyzed in post-processing to, for example, evaluate algebraically the $\Delta u_{k,j}$ quantities needed by multistate analysis (see above).

>>>>

In the alchemical transfer approach, the standard binding free energy is obtained by a coordinate translation that brings the ligand from the solvent to the receptor binding site~\cite{wu2021alchemical} rather than by decoupling it from the system, as in the double-decoupling method. The method derives from Eq.~(\ref{eq:DG0-binding-constant-def}) by inserting the solvent degrees of freedom similarly to the double-decoupling method, except that the product $z_{R(v)} z_{L(v)}$ of partition functions of the unbound state is represented collectively as one explicit solvent system in which the ligand is placed in a region spanned by an indicator function $I^\ast(\zeta)$ identical to the one used to define the binding site region but placed at some arbitrarily large distance from the receptor:
\begin{equation}
  z_{R(v)} z_{L(v)} =
  \frac{
    \int d\vec{r}_v dx_R dx_L d\zeta I^\ast(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
  }{
    8 \pi^2 V_{\rm site} Z_{N_v}
  }
\end{equation}
where, as above, we multiplied and divided by the volume of the binding site, assuming for simplicity that the indicator function does not depend on the orientational degrees of freedom of the ligand. Combining this result with Eq.~(\ref{eq:partfunct-ratio-explsolv-1}) for the complex and canceling the common partition function of the pure solvent, leads with Eq.~(\ref{eq:DG0-binding-constant-def})  to an expression for the binding constant in terms of ratios of partition functions in explicit solvent that differ only by the location of the indicator function (in the binding site or in the solvent region):
\begin{equation}\label{eq:atm-partfunc-ratio}
  K_b = C^\circ V_{\rm site} \frac{
    \int d\vec{r}_v dx_R dx_L d\zeta I(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
  }{
    \int d\vec{r}_v dx_R dx_L d\zeta I^\ast(\zeta) e^{-\beta U(x_R, x_L, \zeta, \vec{r}_v)}
  }
\end{equation}
The idea of alchemical transfer, illustrated in Figure \ref{fig:atm-illustration}A, is to cast Eq.~(\ref{eq:atm-partfunc-ratio}) as an ensemble average by a change of coordinates noting that $I(\zeta) = I^\ast(\zeta - \vec{h})$ where $\vec{h}$ is the fixed vector distance between the binding site and solvent locations spanned by the two indicator functions. The symbol $\zeta$ represents the position and orientation of the ligand relative to the receptor, so here $\zeta - \vec{h}$ is the operation of shifting the position of the ligand by the displacement $\vec{h}$ without changing its orientation. Hence, by applying the change of variable $\zeta \rightarrow \zeta - \vec{h}$ to the integral in the numerator of Eq.~(\ref{eq:atm-partfunc-ratio}), we arrive at the following expression for the binding constant, equivalent to Eq.~(\ref{eq:DG0-binding-constant-def-average-noOmega}):
\begin{equation}\label{eq:atm-average-abfe}
K_b = C^\circ V_{\rm site} \langle e^{-\beta \Delta U} \rangle_0
\end{equation}
where
\begin{equation}\label{eq:atm-deltaU-def}
\Delta U(x) = U(x_R, x_L, \zeta + \vec{h}, \vec{r}_v) - U(x_R, x_L, \zeta, \vec{r}_v)
\end{equation}
is the perturbation energy of the method, and the ensemble average $\langle \ldots \rangle_0$ is over the unbound ensemble when the ligand is limited to the solvent region far away from the receptor. Eq.~(\ref{eq:atm-average-abfe}) instructs to simulate the unbounded system where the ligand is limited to a region in the solvent and to sample the change of potential energy $\Delta U$ resulting from transferring the ligand from the solvent to the receptor binding site by a fixed displacement of the ligand as a whole.

@@
